Value-Suppressing Uncertainty Maps

### 2017-09-19 paper/main.tex
Last minute edits

When selecting a visual channel to represent uncertainty, we recommend channels with both a strong semiotic connection to uncertainty (such as those recommended by MacEachren et al.~\cite{maceachren2012visual}) as well as a relatively large number of perceptually distinguishable levels (such as those evaluated by Boukhelifa et al. \cite{boukhelifa2012evaluating}). Here, we focus on a combination of increasing color value and decreasing saturation. Correll \& Gleicher~\cite{correll2013error} provide evidence that even audiences without statistical training can successfully interpret uncertainty information encoded in these channels.

@@

Next generation sequencing allows biologists to collect and analyze large amounts of genomics data. However, these techniques often create data quality issues, as they must align large numbers of potentially ambiguous ``reads'' of relatively small numbers of base pairs. Using a dataset from a group of virologists interested in the population dynamics of viral mutation~\cite{o2012conditional}, we visualized the variability in viral genomes across 25 populations of the simian immunodeficiency virus (SIV). In the dataset here, it is important not only to identify hotspots in the viral genome (locations with high rates of mutation or variability), but also to not be distracted by false positives. The traditional bivariate map (\figref{fig:viral2d}) has many locations with apparent inter-population variance.

@@

We limited our population to Turkers from the United States, with a prior task approval rate of at least 90\%. As the experimental tasks required multi-hue color perception, we presented participants with a set of Ishihara plates~\cite{hardy1945tests} as a pre-test, and excluded participants who either misidentified the values in the plates, or who self-reported as having a color vision deficiency (CVD) in the post-test. Based on piloting, we paid participants \$2 dollars for participation, for a target rate of \$8/hour. After completion of the main tasks, we solicited demographic information, including a risk aversion assay from Mandrik \& Bao~\cite{mandrik2005exploring}. Participants from the first experiment were excluded from participating in the second.

@@

The results support our first hypothesis. Juxtaposition was a significant effect ($F(1,166)=5.5$,$p=0.02$). A post-hoc t-test confirmed ($t=2.5$, $p=0.01$) that superimposed charts ($M=0.58$, $SD=0.5$) performed significantly better than juxtaposed ($M=0.51$, $SD=0.49$) ones. For our identification task, juxtaposition requires at least two search tasks (one to locate squares with the correct data value, another to locate ones with the correct uncertainty value), which introduces additional error compared to conditions where value and uncertainty are displayed simultaneously. This is reflected also in response times, which, although highly variable, were highest in juxtaposed conditions ($M=19$s, $SD=65$), than the other conditions ($M=13$s, $SD=38$), although this difference was not significant ($t=1.64$, $p=0.1$).

@@

We performed a second ANOVA among the superimposed discrete charts to determine the effect of legend shape (wedge or square) and quantization scheme (VSUP or standard) on performance, with participant ID as a random factor. We did not find a significant effect for either the legend shape ($F(1,70)=0.03$, $p=0.084$) or the quantization scheme ($F(1,70)=1.4$, $p=0.24$).

>>>>

When selecting a visual channel to represent uncertainty, we recommend channels with both a strong semiotic connection to uncertainty (such as those recommended by MacEachren et al.~\cite{maceachren2012visual}) as well as a relatively large number of perceptually distinguishable levels (such as those evaluated by Boukhelifa et al. \cite{boukhelifa2012evaluating}). Here, we focus on a combination of increasing luminance and decreasing saturation. Correll \& Gleicher~\cite{correll2013error} provide evidence that even audiences without statistical training can successfully interpret uncertainty information encoded in these channels.

@@

Next generation sequencing allows biologists to collect and analyze large amounts of genomics data. However, these techniques often create data quality issues, as they must align large numbers of potentially ambiguous ``reads'' of relatively small numbers of base pairs. Using a dataset from a group of virologists interested in the population dynamics of viral mutation~\cite{o2012conditional}, we visualized the variability in viral genomes across 12 populations of the simian immunodeficiency virus (SIV). In the dataset here, it is important not only to identify hotspots in the viral genome (locations with high rates of mutation or variability), but also to not be distracted by false positives. The traditional bivariate map (\figref{fig:viral2d}) has many locations with apparent inter-population variance.

@@

We limited our population to Turkers from the United States, with a prior task approval rate of at least 90\%. As the experimental tasks required multi-hue color perception, we presented participants with a set of Ishihara plates~\cite{hardy1945tests} as a pre-test, and excluded participants who either misidentified the values in the plates, or who self-reported as having a color vision deficiency (CVD) in the post-test. Based on piloting, we paid participants \$2 for participation, for a target rate of \$8/hour. After completion of the main tasks, we solicited demographic information, including a risk aversion assay from Mandrik \& Bao~\cite{mandrik2005exploring}. Participants from the first experiment were excluded from participating in the second.

@@

The results support our first hypothesis. Juxtaposition was a significant effect ($F(1,166)=5.5$,$p=0.02$). A post-hoc t-test confirmed ($t=2.5$, $p=0.01$) that superimposed charts ($M=0.58$, $SD=0.5$) performed significantly better than juxtaposed ($M=0.51$, $SD=0.49$) ones. For our identification task, juxtaposition requires at least two search tasks (one to locate squares with the correct data value, another to locate ones with the correct uncertainty value), which introduces additional error compared to conditions where value and uncertainty are displayed simultaneously. 
%This is reflected also in response times, which, although highly variable, were higher in juxtaposed conditions ($M=19$s, $SD=65$), than the other conditions ($M=13$s, $SD=38$), although this difference was not significant ($t=1.64$, $p=0.1$).

@@

We performed a second ANOVA among the superimposed discrete charts to determine the effect of legend shape (wedge or square) and quantization scheme (VSUP or standard) on performance, with participant ID as a random factor. We did not find a significant effect for either the legend shape ($F(1,70)=0.04$, $p=0.84$) or the quantization scheme ($F(1,70)=1.4$, $p=0.24$).
